---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# IDmeasurer
The goal of `IDmeasurer` package is to provide tools for assessment and quantification of individual identity information in animal signals. This package accompanies a research article by Linhart et al.: ['Measuring individual identity information in animal signals: Overview and performance of available identity metrics'](https://www.biorxiv.org/content/10.1101/546143v1), which can currently be accessed at BioRxive.  

## Installation
The package is currently available at GitHub:

``` r
devtools::install_github('pygmy83/IDmeasurer', build = TRUE, build_opts = c("--no-resave-data", "--no-manual"))
```

The package has been also submitted to CRAN and it should be soon possible to install the released version of `IDmeasurer` from [CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("IDmeasurer")
```

## Example

This is a basic example which shows how to calculate individual identity information in territorial calls of little owls (`ANspec` example data):

```{r}
library(IDmeasurer)
```

Input data for the calculation of identity metrics in this package, in general, is a data frame with the first column containing individual identity codes (factor) and the other columns containing individuality traits (numeric).

```{r}
summary(ANspec)   
```

This calculates HS metric for every single trait variable in the dataset.
```{r}
calcHS(ANspec, sumHS=F)
```

To calculate the HS for an entire signal, it is neccessary to have uncorrelated variables in dataset. Raw (correlated) trait variables need to be transformed into principal components by the Principal component analysis.
```{r}
temp <- calcPCA(ANspec) 
```

Calculate HS for an entire signal.
```{r}
calcHS(temp) 
```

To see description of the example dataset, use:

``` r
?ANspec
```

More examples can be found in IDmeasurer vignette:
``` r
vignette('idmeasurer-workflow-examples')
```

